---
title: "Practicum I CS5200"
author: "Wenhao Li"
date: "Summer Full 2023"
output:
  pdf_document: default
  html_notebook: default
email: li.wenhao3@northeastern.edu
---

## Connect to Database

Establish link to database at freemysqlhosting.net
```{r}
library(RMySQL)

db_name_fh <- "sql3654492"
db_user_fh <- "sql3654492"
db_host_fh <- "sql3.freemysqlhosting.net"
db_pwd_fh <- "LfRIZIav7w"
db_port_fh <- 3306

mydb.fh <-  dbConnect(RMySQL::MySQL(), user = db_user_fh, password = db_pwd_fh,
                      dbname = db_name_fh, host = db_host_fh, port = db_port_fh)

dbcon <- mydb.fh
```

## Create Database

Table for airport
```{sql connection=dbcon}
CREATE TABLE IF NOT EXISTS airport(
  aid INTEGER AUTO_INCREMENT PRIMARY KEY,
  airportName VARCHAR(50) DEFAULT 'unknown',
  airportState VARCHAR(50) DEFAULT 'unknown',
  airportCode VARCHAR(50) DEFAULT 'unknown'
)
```

Table for each flights
```{sql connection=dbcon}
CREATE TABLE IF NOT EXISTS flights(
  fid INT PRIMARY KEY,
  flight_date DATE DEFAULT NULL,
  origin INT DEFAULT 1,
  airline VARCHAR(50) DEFAULT "unknown",
  aircraft VARCHAR(50) DEFAULT "unknown" ,
  altitude INT UNSIGNED DEFAULT 0,
  heavy BOOLEAN DEFAULT FALSE,
  CONSTRAINT flights_fk_a FOREIGN KEY(origin) REFERENCES airport(aid)
)
```

Look up table: conditions
```{sql connection=dbcon}
CREATE TABLE IF NOT EXISTS conditions(
  cid INT AUTO_INCREMENT PRIMARY KEY,
  sky_condition VARCHAR(50) DEFAULT "unknown",
  explanation VARCHAR(50) DEFAULT ""
)
```

Create table: strike
```{sql connection=dbcon}
CREATE TABLE IF NOT EXISTS strike(
  sid INT AUTO_INCREMENT PRIMARY KEY,
  fid INT NOT NULL DEFAULT 1,
  numbirds INT UNSIGNED DEFAULT 0,
  impact VARCHAR(50) DEFAULT "unknown",
  damage BOOLEAN DEFAULT FALSE,
  altitude INT UNSIGNED DEFAULT 0,
  conditions INT DEFAULT 1,
  CONSTRAINT positive_altitude CHECK(altitude >= 0),
  CONSTRAINT strike_fk_f FOREIGN KEY(fid) REFERENCES flights(fid),
  CONSTRAINT strike_fk_c FOREIGN KEY(conditions) REFERENCES conditions(cid)
)
```

## Tests for table implantation

```{sql connection=dbcon, eval = F}
INSERT INTO airport VALUES
  ()
```


Should fail because of foreign key constrain
```{sql connection=dbcon, eval = F}
INSERT INTO flights (fid, flight_date, origin, airline, aircraft, altitude, heavy) VALUES
  (202152, "2000-11-23", 1, "US AIRWAYS*", "B-737-400", 1500, TRUE)
```
Tests for table implantation
```{sql connection=dbcon, eval = F}
INSERT INTO airport (airportName, airportState, airportCode) VALUES
  ("LAGUARDIA NY", "New York", "1"),
  ("DALLAS/FORT WORTH INTL ARPT", "Texas", "2")
```
Tests for table implantation
```{sql connection=dbcon, eval = F}
SELECT *
FROM airport
```
altitude should convert to default value
```{sql connection=dbcon, eval = F}
INSERT INTO flights 
SET fid = 202152,
    flight_date = "2000-11-23",
    airline = "US AIRWAYS*",
    aircraft = "B-737-400",
    altitude = -1500,
    heavy = TRUE,
    origin = (
    SELECT aid
    FROM airport
    WHERE airportName = "LAGUARDIA NY")
```
Tests for table implantation
```{sql connection=dbcon, eval = F}
SELECT * FROM flights
```
Tests for table implantation
```{sql connection=dbcon, eval = F}
INSERT INTO conditions (sky_condition, explanation) VALUES
  ("NO CLOUD", ""),
  ("Some Cloud", "")
```
Tests for table implantation
```{sql connection=dbcon, eval = F}
SELECT * FROM conditions
```
Tests for table implantation
```{sql connection=dbcon, eval = F}
INSERT INTO strike 
SET numbirds = 859,
    impact = "ENGINE SHUT DOWN",
    damage = TRUE,
    altitude = -1500,
    fid = 202152,
    conditions = 1
```
Tests for table implantation
```{sql connection=dbcon, eval = F}
SELECT * FROM strike
```
Tests for table implantation
```{sql connection=dbcon, eval = F}
DELETE FROM airport
```
Tests for table implantation
```{sql connection=dbcon, eval = F}
DELETE FROM flights
```
Tests for table implantation
```{sql connection=dbcon, eval = F}
DELETE FROM conditions
```
Tests for table implantation
```{sql connection=dbcon, eval = F}
DELETE FROM strike
```

## Populate Database

Adding Default Conditions for each table except strike
```{sql connection=dbcon}
INSERT INTO airport (airportCode) VALUES
  ("DEFAULT")
```

Adding Default Conditions for each table except strike
```{sql connection=dbcon}
INSERT INTO flights
SET fid = 0,
    origin = (SELECT aid
              FROM airport
              WHERE airportName = "unknown" AND airportCode = "DEFAULT")
```

Adding Default Conditions for each table except strike
```{sql connection=dbcon}
INSERT INTO conditions VALUES
  ()
```

Installing and loading of libraries
```{r, warning=FALSE}
install.packages("dplyr")
library(dplyr)
```
Manipulate dataframe for populating table
```{r}
fileName <- "BirdStrikesData-V2.csv"

df <- read.csv(fileName)

df <- data.frame(lapply(df, function(x) {gsub("\\*", "", x)}))

#Add "" to each column which will be inserted into the database as TEXT
df$airport <- paste('"', df$airport, '"')
df$model <- paste('"', df$model, '"')
df$impact <- paste('"', df$impact, '"')
df$airline <- paste('"', df$airline, '"')
df$origin <- paste('"', df$origin, '"')
df$sky_conditions <- paste('"', df$sky_conditions, '"')

#Change all colume that is empty or "N/A" to DEFAULT ("unknown" for TEXT, FALSE for boolean, 0 for INT)
df <- data.frame(lapply(df, function(x) {gsub('"  "', "DEFAULT", x)}))
df <- data.frame(lapply(df, function(x) {gsub('" N/A "', "DEFAULT", x)}))

#Delete space in TEXT
df <- data.frame(lapply(df, function(x) {gsub('" ', '"', x)}))
df <- data.frame(lapply(df, function(x) {gsub(' "', '"', x)}))
```


Importing Data into table airport
```{r}
#Group by function delete repeating rows
airport_df <-  df[] %>% group_by(airport, origin) %>%
                    summarise(.groups = 'drop',
                              empty_column = "",)

#Run insert for each row in airport_df
for (i in 1:nrow(airport_df)){
  row <- airport_df[i,]
  sql <- paste0('INSERT INTO airport (airportName, airportState, airportCode) VALUES (', 
                row[1], ', ', row[2], ', DEFAULT)')
  dbExecute(dbcon, sql)
}
```

Importing Data into conditions
```{r}
#Group by function delete repeating rows
conditions_df <- df[] %>% group_by(sky_conditions) %>%
                    summarise(.groups = 'drop',
                              empty_column = "",)

#Run insert for each row in conditions_df
for (i in 1:nrow(conditions_df)){
  row <- conditions_df[i,]
  sql <- paste0('INSERT INTO conditions (sky_condition, explanation) VALUES (', 
                row[1], ', ', row[2], 'DEFAULT)')
  dbExecute(dbcon, sql)
}
```

Importing Data into table flights
```{r}
#Group by function delete repeating rows
flights_df <- df[] %>% group_by(rid, flight_date, airport, origin, airline, model, altitude_ft, heavy_flag) %>%
                    summarise(.groups = 'drop')

#Convert date column to acceptable format
flights_df$flight_date <- as.Date(flights_df$flight_date, "%m/%d/%Y")
flights_df$flight_date <- as.character(flights_df$flight_date)

#Convert boolean columns to acceptable format
flights_df["heavy_flag"][flights_df["heavy_flag"] != "Yes"] <- "FALSE"
flights_df["heavy_flag"][flights_df["heavy_flag"] == "Yes"] <- "TRUE"

#Convert Default to coresponse value for sql query in FK insertion
flights_df["airport"][flights_df["airport"] == "DEFAULT"] <- '"unknown"'
flights_df["origin"][flights_df["origin"] == "DEFAULT"] <- '"unknown"'
flights_df["altitude_ft"][flights_df["altitude_ft"] == ""] <- "0"
flights_df["flight_date"][is.na(flights_df["flight_date"])] <- "DEFAULT"

#Delete comma in column altitude
flights_df$altitude_ft <- gsub(",","",flights_df$altitude_ft)

#Run insert for each row in flights_df
for (i in 1:nrow(flights_df)){
  row <- flights_df[i,]
  sql <- paste0('INSERT INTO flights SET fid = ', 
                row[1], ', flight_date = "', row[2], 
                '", origin = (SELECT aid FROM airport WHERE airportName = ',
                row[3], 'AND airportState = ', row[4],'), airline = ', row[5], 
                ', aircraft = ', row[6], ', altitude = ', row[7], ', heavy = ', row[8])
  dbExecute(dbcon, sql)
}
```

Importing Data into table strike
```{r}
#Group by function delete repeating rows
strike_df <- df[] %>% group_by(rid, wildlife_struck, impact, damage, altitude_ft, sky_conditions) %>%
                    summarise(.groups = 'drop')

#Convert boolean columns to acceptable format
strike_df["damage"][strike_df["damage"] != "No damage"] <- "TRUE"
strike_df["damage"][strike_df["damage"] == "No damage"] <- "FALSE"
strike_df["altitude_ft"][strike_df["altitude_ft"] == ""] <- "0"

#Delete comma in column altitude
strike_df <- data.frame(lapply(strike_df, function(x) {gsub("\\,", "", x)}))

#Run insert for each row in strike_df
for (i in 1:nrow(strike_df)){
  row <- strike_df[i,]
  sql <- paste0('INSERT INTO strike SET fid = (SELECT fid FROM flights WHERE flights.fid = ', 
                row[1], '), conditions = (SELECT cid FROM conditions WHERE sky_condition = ', row[6],
                '), numbirds = ', row[2], ', impact = ', row[3], ', damage = ', row[4], 
                ', altitude = ', row[5])
  dbExecute(dbcon, sql)
}
```

Check table airport
```{sql connection=dbcon}
SELECT *
FROM airport
LIMIT 0,10
```

Check table conditions
```{sql connection=dbcon}
SELECT *
FROM conditions
```

Check table flights
```{sql connection=dbcon}
SELECT *
FROM flights
LIMIT 0, 10
```

Check table strike
```{sql connection=dbcon}
SELECT *
FROM strike
LIMIT 0, 10
```

## DATA ANALYSIS

Find the top 10 states with the greatest number of bird strike incidents
```{sql connection=dbcon}
SELECT airportState AS state, COUNT(*) AS incidentsNum
FROM strike s
  INNER JOIN flights f ON s.fid = f.fid
  INNER JOIN airport a ON f.origin = a.aid
GROUP BY a.airportState
HAVING a.airportState != "unknown"
ORDER BY incidentsNum DESC
LIMIT 10
```

Find the airlines that had an above average number bird strike incidents
```{sql connection=dbcon}
SELECT airline, incidentsNum
FROM (SELECT f.airline, COUNT(*) AS incidentsNum
      FROM strike s
        INNER JOIN flights f ON s.fid = f.fid
      GROUP BY f.airline
      HAVING f.airline != "unknown") AS t
WHERE incidentsNum > 
      (SELECT AVG(incidentsNum) AS Average
      FROM (SELECT f.airline, COUNT(*) AS incidentsNum
            FROM strike s
              INNER JOIN flights f ON s.fid = f.fid
            GROUP BY f.airline
            HAVING f.airline != "unknown") AS t)
```

find the (total) number of birds that struck aircraft by month
```{r}
sql = "SELECT EXTRACT(MONTH FROM f.flight_date) AS month, COUNT(*) AS incidentsNum
       FROM flights f
         INNER JOIN strike s ON s.fid = f.fid
       WHERE f.flight_date != '0000-00-00'
       GROUP BY month"

month_df <- dbGetQuery(dbcon, sql)

print(month_df)
```

plot bird strike distribution each month
```{r}
barplot(month_df$incidentsNum, beside = FALSE, main = "Bird Strike Distribution Each Month",
        names.arg = month_df$month, xlab = "Month", ylab = "number of incidents")
```

## Store Procedure for add new incidents

adds a new strike to the database (store procedure)
```{sql connection=dbcon}
CREATE PROCEDURE newEntry (IN new_sky_condition TEXT, IN new_airportName TEXT, 
                           IN new_airportState TEXT, IN new_fid INT, IN new_flight_date DATE,
                           IN new_airline TEXT, IN new_aircraft TEXT, IN new_altitude INT,
                           IN new_heavy BOOLEAN, IN new_numbirds INT, IN new_impact TEXT,
                           IN new_damage BOOLEAN)
BEGIN
  -- check if sky_condition in new entry have store in database, if not, enter it into database
  IF (!EXISTS (SELECT * FROM conditions WHERE sky_condition = new_sky_condition)) THEN
    INSERT INTO conditions (sky_condition, explanation) VALUES (new_sky_condition, "");
  END IF;
  -- check if airport in new entry have store in database, if not, enter it into database
  IF (!EXISTS (SELECT * FROM airport WHERE airportName = new_airportName 
                                       AND airportState = new_airportState)) THEN
    INSERT INTO airport (airportName, airportState) VALUES (new_airportName, new_airportState);
  END IF;
  -- check if flight in new entry have store in database, if not, enter it into database
  IF (!EXISTS (SELECT * FROM flights WHERE fid = new_fid)) THEN
    INSERT INTO flights SET fid = new_fid, 
                            flight_date = new_flight_date,
                            origin = (SELECT aid FROM airport WHERE airportName = new_airportName
                                                                AND airportState = new_airportState),
                            airline = new_airline,
                            aircraft = new_aircraft,
                            altitude = new_altitude,
                            heavy = new_heavy;
  END IF;
  -- add new insidents
  INSERT INTO strike SET fid = (SELECT fid FROM flights WHERE fid = new_fid),
                         numbirds = new_numbirds,
                         impact = new_impact,
                         damage = new_damage,
                         altitude = new_altitude,
                         conditions = (SELECT cid FROM conditions 
                                       WHERE sky_condition = new_sky_condition);
END
```
Test Store Procedure
```{sql connection=dbcon}
CALL newEntry("OK", "OK", "OK", 11, "1000-10-10", "OK", "OK", 0, FALSE, 1, "OK", FALSE)
```
Test Store Procedure
```{sql connection=dbcon}
SELECT *
FROM strike
WHERE fid = 11
```
Test Store Procedure
```{sql connection=dbcon}
DELETE FROM strike WHERE fid = 11
```
Test Store Procedure
```{sql connection=dbcon}
DELETE FROM flights WHERE fid = 11
```
Test Store Procedure
```{sql connection=dbcon}
DELETE FROM conditions WHERE sky_condition = "OK"
```
Test Store Procedure
```{sql connection=dbcon}
DELETE FROM airport WHERE airportName = "OK"
```